CSD, Disjoint components and trayectory detection

### 2017-11-08 ArticuloCSDyCM/Paper03.tex
cambios menores

Spatially tracking the activity of electro-physiological data in a concise, rigorous and interpretable manner poses a challenging problem. Measurement recordings are usually presented first as a series of traces that correspond each to a sampling point on the tissue. As Micro Electrode Arrays (MEAs) become denser, this representation of the data becomes saturated and the discovery of patterns turns into a difficult task. Novel information is obfuscated by the same high density acquisition that allows us to discover it. Representing  the recorded data in ways which can unveil previously obfuscated information is a necessity. Various techniques for separation of functional generators are available now, most come from the application of blind source separation analysis, such as ICA. 
Our task in this paper is to present a method for tracing out ``functional connectivity'', which would bring independent information that is complementary to those analysis.  Our  method would be applicable mostly to activity in structured neural tissue. We shall exemplify it by application to the data obtained from recordings of the rats hippocampus.

@@

In previous works some form of vector averaging has been proposed and shown to reveal some trajectrories drawn by the activity. We shall call such methods Center of Mass Analyses (CMA).  The \emph{wheights} that have been chosen for obtaining the average have been selected on heuristic grounds. In the works of Chao \emph{et al.} \cite{Chao05, Chao07}, the density of spikes in a certain time interval was used, a sort of ``density of action'', while Manjarrez \emph{et al.} \cite{Manjarrez07, Manjarrez09} used the positive half of the LFP recording as their weight. 
Here we advance on such methods by incorporating mathematically sound definitions and concepts which allow us to rigorously separate spatially distinct traces. Our weight is actually the natural density function provided by the data: the Current Source Density. This representation not only provides a mathematical rigorous  measure for the weights in the vector averaging process, but helps to separate the data into disjoint components. This allows us to track separate process in the same region of interest.

The application of the method requires electro-physiological data acquired through bi-dimensional MEAs, although the extension to three dimensional arrays would, in principle, be possible. The MEA should have dense enough spatial and temporal samplings, so as to obtain good estimates of the CSD and their changes in time.

@@

The concepts presented here require first obtaining the CSD from the recorded LFP. Any correct method would present the same utility. The first method is a  finite difference operator, while the second is the kCSDA presented by Potworowski \emph{et all} \cite{Potworowski2011}, which is an inverse problem approach.  Where both methods are usable, we prefer the former, as is faster and requires less assumptions. We present to methods and show the robustness of the result for the Center of Mass tracking.

@@

On previous works it has been argued that numerical difference operators sacrifice all data on the borders of the array. For electrode arrays of lesser density, this could posse a problem, but in our case we only renounce to 27 out of 4095 channels (one is grounded), and those are outside the region of interest.
If very few electrodes fail, more sophisticated methods such as iCSDA \cite{Leski2011} and kCSDA \cite{Potworowski2011} prove little advantage in estimating the sources under the ohmic, isotropic and homogeneous assumptions. In case that around 30 or more electrodes
fail, specially if they are on the region of interest, the use of such methods
becomes necessary. The kCSDA method relies on the
assumption that a finite family of reasonable functions can span (with acceptable accuracy) the possible CSD at each instant in time. The idea behind the method is simple:
presenting one model function of this family and obtaining the LFP that such a function
would generate, we calculate the coefficients for the entire family by projecting the
experimental data over this ideal LFP model function family. A general presentation is in the reference, here
we shall recount only the implementation that we used.

We shall borrow the notation in \cite{Potworowski2011}. Our model function for the CSD is a hard disk function, given by:
\begin{equation}
  \tilde{b}_j(x,y)=\begin{cases}
  1, \text{ for } (x-x_j)^2+(y-y_j)^2 \leq R^2 \\
  0 \text{ otherwise.} 
  \end{cases}
\end{equation}
The $R$ parameter is the effective radius of the charge density, which for our
calculations was taken as half the inter-electrode distance, that is $R=21 \mu m$.
Each pair $(x_j, y_j)$ is the center of one such model function, which we take as
the position of each  functioning electrode in an adequate coordinate system.
The LFP generated by this function would then be:
\begin{equation}
  \tilde{b}_j(x,y)=\frac{1}{2\pi\sigma}
  \iint \! \rd x \rd y \arcsinh \biggr(
  \frac{2 h } {\sqrt{(x-x_j)^2+(y-y_j)^2}}
  \biggl)
  \tilde{b}_j(x,y),
\end{equation}
  with $h$ being the effective height of the function support, in our case,
  the half width of the hippocampus slice ($h=150 \mu m$).

@@

It is important to mention that our data has noise up to 0.1 of the magnitude of the
signal, which means that regularization of the matrix $K$ is unavoidable (see section 4.1 in \cite{Potworowski2011} and references therein). We used a regularization parameter $\lambda=0.001 |e_{max}|$, where $e_{max}$ is the principal eigenvalue of $K$.

The attainment of $K$ and $\tilde{K}$ turn out to be computer-costly for high density MEAs and dense temporal sampling, but paraletization, specially on
GPUs, solves this issue.

We have performed comparison of both methods over experimental data that has
no failing channels in order to adjust the parameters of the second method. We
have obtained consistent results between them.

@@

Center of Mass Analysis (CMA) has been used previously in order to track putative  ``trajectories'' of the activity or information across neural tissue or even the whole brain \cite{Chao05, Chao07, Manjarrez07, Manjarrez09}. In such works the chosen ``mass'' was an heuristic measure adecuate for representing the corse-grain picture of displacement of activity. These present the problem of justifing the choice made on practical grounds. For calculating a vectorial average, the adecuate wheight per vector is a density, a positive quantity that measures the concentration of something per space unit, so it is only adequate that we search for a natural density in the electrophysiological experiments. The Current Source Density fits the mathematical and interpretative requirements. We shall denote it in the following as $C$ (following \cite{Potworowski2011}).  This measure is usually presented in arbitrary units, which permits us to interpret it two-fold. One, directly from its definition, as the \emph{differential divergence} of the vectorial electric current density:
\begin{equation}
  C:=\nabla \cdot J
\end{equation}
Due to conservation of charges, this is the same as as the temporal variation of density of charge:
\begin{equation}
  C=\frac{\partial \rho}{\partial t}
\end{equation}
Current Source Density as a mathematical density (a quasi-probability density function)  allows  to rigorously apply the concepts of vector averaging. This use has been overlooked. This might have occurred because  samplings at lesser spatial resolution did not permit estimations of this quantity with enough precision to make reliable spatial implications, even with the use of sophisticated CSD estimation methods. In our case, we want to use the density as a mathematician would: a function which assigns weights to points in the space of interest and permits us to carry linear operations over them.  The space here is the sampling points of the electrodes at a given instant in time, and the weight, the CSD, is amount interchange of charged ions from the inside to the outside of cellular membranes, per unit space.  This is an indicator of activity in the neurons, therefore the CSD is also, indirectly, a \emph{``density of local activity''}.

Apparently, CSD would not fit the requirements of non-negativity for its use as a mathematical density, but this turns out to be not a problem.  CSD separates the measured activity into three distinct sets, namely, the set of all sources, the set of all sinks and the zero set. The separation is not quantitative but qualitative: even when the signs of electric charge are arbitrary, the distinction between them is not. Therefore, CSD represents on each of the two active sets (sinks and sources) exactly what we need, a unit of concentration per unit space, in this case, per area, and the sign can be ignored, as long as we perform the operations in the sets separately. Also, the separation by the zero value set of the CSD is in contrast to the zero value of the LFP, where it has no precise meaning as it only represents a practical reference value. This is acknowledged in most of the literature, where only the scales of the LFP recording are shown, without reference to the sign of the values. In the CSD representation,  the so called zero set (all the points in our space which have a recorded CSD of zero)  has a precise meaning: it is an instantaneous lack of activity and a border between sinks and sources. This shall be exploited further in the section below.

@@

To clarify the concept and how we are we using it an example is in order. In figure \ref{Disjuntos} A we show geometrical sets enclosed by simple curves as color patches, where every color indicates a Set. The set $A$ consists of a single component, that means that one can make  path between any two points in the set that consist entirely of points of the same set. In contrast, the set $B$ is made of two disjoint components, labeled $B1$ and $B2$. If one chooses a point in each of the components, one cannot draw a path joining them without leaving the set. But each one of those components is connected, meaning that if we choose the pair of points in one component, then again we can trace a path joining them in the set. So the set $B$ consists of two disjoint components, but each component can be regarded as a connected set on their own. Likewise the set $C$ is made of three disjoint components. In the subfigre \ref{Disjuntos} Aiii we show the center of mass of the sets. The set $A$ is concave, so, it can have a center of mass outside the set. The
set $B$ has also a center of mass outside the set. It has two disjoint components that are aproximatelly
equall and separated by a long distance (this is not an implication: a set can be disjoint and have
a CM inside the set). For the set $C$ our aproach has been used: we treat every disjoint component of the
set as an independent entity and calculate the center of mass for each one. Notice that this doesn't preclude the posibility of having a concave subset with a CM that lies outside. But, as we shall see,
as the electrode resolution is finite, the posibility of having concave shapes is unlikely, and they
tend to have a CM near their borders. In the figure \ref{Disjuntos} B we show this with experimental
data (we are using evoked activity data, as explained in section \ref{sec:evocada}). In graph Bi we
sepparate the CSD data into the sink and sources set. We take sinks as example. In the grabh Bii
we label each disjoint component, obtained by a single pass algorhythm. There are 25 components, and
the biggest two are not strictly convex but they have tolerable deviations from it. In the
subfigure Biii we indicate the CM of only those components that are above certain threshold of
integrated intensity (the dot is proportional in area to such intensity, there are five CM to be
seen). Each CM gets listed as a probable step in a trajectory.

\begin{figure}
  \includegraphics[width=0.9\textwidth]{Figuras/Disjuntos01.pdf}
  \caption{ Explanation of the geometrical concepts used on our analysis. In subfigure A we show
    with an artificial example the concepts of disjointness and convexivity. The subfigure Ai shows
    three sets, $A$, $B$, and $C$. $A$ fails to be convex because it is concave, and $C$ fails to be
    be convex because it is made of two disjoint components. In each of them is possible to draw a line
    connecting two points inside the set that is not entirely inside the set. In $B$ that is not possible, the set $B$ is convex in a mathematical sense. In subfigure Aii we made some modification to whos that
    each set can be made of many disjoint components, and each one of them can be convex or not.  In Aiii we show the center of mass for $A$ and $B$ taken as a wholle, while for $C$ we break the set into each of the disjoint components and calculate each CM separately. There is no guarantee that each disjoint component would be convex, but as the resolution of the data is finite, the smaller components are more likely to
    be convex. In figure B we shoe everything with a frame of experimental data. In Bi we sepparate the
    data into three sets, Sources, Sinks, and Zeros. We discard the Zeros and keep the interesting ones.
    In Figure Bii we take the Sink Set and separate it into disjoint components, that are labeled from one to 25 in this particular example. From this subsets we calculate the total mass of each one and only
    considerate for the next step those who are above certain threshold of integrated intensity. The
    next step is shown in Biii, where we plot the Center of Mass for each of such subsets. Deviations
    from convexity seem to be unimportant, as most of the mass is concentrated in the inner electrodes
    of each set.
  }
  \label{disjuntos}
\end{figure}

@@

We must not take out of sight the phenomena which is producing the CSD distribution: neurons have a finite size, and the effect of their action a delimited extent. Every patch that we can identify as a disjoint component of the Sources/Sink sets indicates a coherent, joint active group of units. A center of mass for each disjoint component would then indicate an average locus for this active group, a putative center of activity to which we can give a position, intensity and time. An instantaneous snapshot of such information would depict the centers of activity of an ordered structure of neurons and their relative intensities. But it is in the successive depiction of a series of such snapshots that this analysis can show interpretative power. The appearance , displacement, and disappearance of such centers could reveal very subtle details of the dynamics of neural activity. Very small, fast displacements of the Centers of mass, may be occur due to slight asynchronous between neurons which belong to the same active group. Larger and  longer than a few firing periods displacements could indicate physiological connections between different groups.

>>>>

Spatially tracking the activity of electro-physiological data in a concise, rigorous and interpretable manner poses a challenging problem. Measurement recordings are usually presented first as a series of traces that correspond each to a sampling point on the tissue. As Micro Electrode Arrays (MEAs) become denser, this representation of the data becomes saturated and the discovery of patterns turns into a difficult task. Novel information is obfuscated by the same high density acquisition that allows us to discover it. Representing  the recorded data in ways which can unveil previously obfuscated information is a necessity. Various techniques for separation of functional generators are available now, most come from the application of blind source separation analysis, such as ICA. 
Our task in this paper is to present a method for tracing out ``functional connectivity'', which would bring independent information that is complementary to those analysis.  Our  method would be applicable mostly to activity in highly structured neural tissue. We shall exemplify it by application to the data obtained from recordings of the rats hippocampus, taking advantage of its lamminar
orderer structure.

@@

In previous works some form of vector averaging has been proposed and shown to reveal some trajectrories drawn by the activity. We shall call such methods Center of Mass Analyses (CMA).  The \emph{wheights} that have been chosen for obtaining the average have been selected on heuristic grounds. In the works of Chao \emph{et al.} \cite{Chao05, Chao07}, the density of spikes in a certain time interval was used, a sort of ``density of action'', while Manjarrez \emph{et al.} \cite{Manjarrez07, Manjarrez09} used the positive half of the LFP recording as their weight. 
Here we advance on such methods by incorporating mathematically sound definitions and concepts which allow us to rigorously separate spatially distinct traces. Our weight is actually the natural density function provided by the data: the Current Source Density. This representation not only provides a mathematical rigorous  measure for the weights in the vector averaging process, but helps to separate the data into disjoint components. This allows us to track separate process in the same region of interest (r.o.i.) by tracking sepparately the poles of action: influx and eflux of charged ions.

The application of the method requires electro-physiological data acquired through bi-dimensional MEAs, although the extension to three dimensional arrays would be possible in principle. The MEA should have dense enough spatial and temporal samplings, so as to obtain good estimates of the CSD and their changes in time.

@@

The concepts presented here require first obtaining the CSD from the recorded LFP. Any method would present the same utility. The first method is a  finite difference operator, while the second is the kCSDA presented by Potworowski \emph{et all} \cite{Potworowski2011}, which is an inverse problem approach.  Where both methods are usable, we prefer the former, as is faster and requires less assumptions. We present both methods and show the robustness of the result for the Center of Mass tracking.

@@

On previous works it has been argued that numerical difference operators sacrifice all data on the borders of the array. For electrode arrays of lesser density, this could posse a problem, but in our case we only renounce to 27 out of 4095 channels (one is grounded), and those are outside the region of interest.
If very few electrodes fail, more sophisticated methods such as iCSDA \cite{Leski2011} and kCSDA \cite{Potworowski2011} prove little advantage in estimating the sources under the ohmic, isotropic and homogeneous assumptions. In case that around 30 or more electrodes
fail, specially if they are on the region of interest, the use of such methods
becomes necessary. The kCSDA method relies on the
assumption that a finite family of convenient functions can span, with acceptable accuracy, the possible CSD at each instant in time. The idea behind the method is simple:
One presents a model function of this family and obtains the LFP that such a function
would generate. Then one calculates the coefficients for the entire family by projecting the
experimental data over this ideal LFP model function family. A general presentation is in the reference \cite{Potworowski2011},we shall recount only the basic implementation that we used.

We shall borrow the notation in the original paper \cite{Potworowski2011}. Our model function for the CSD is a hard disk function, described by:
\begin{equation}
  \tilde{b}_j(x,y)=\begin{cases}
  1, \text{ for } (x-x_j)^2+(y-y_j)^2 \leq R^2 \\
  0 \text{ otherwise.} 
  \end{cases}
\end{equation}
The $R$ parameter is the effective radius of the charge density, which for our
calculations was taken as half the inter-electrode distance, that is $R=21 \mu m$.
Each pair $(x_j, y_j)$ is the center of one such model function, which we take as
the position of each  functioning electrode in an adequate coordinate system.
The LFP generated by this function would then be:
\begin{equation}
  \tilde{b}_j(x,y)=\frac{1}{2\pi\sigma}
  \iint \! \rd x \rd y \arcsinh \biggr(
  \frac{2 h } {\sqrt{(x-x_j)^2+(y-y_j)^2}}
  \biggl)
  \tilde{b}_j(x,y),
\end{equation}
  with $h$ being the effective height of the function support, in our case,
  the half width of the hippocampus slice ($h=150 \mu m$).

@@

It is important to mention that our data has noise up to 0.1 of the mean magnitude of the
signal, which means that regularization of the matrix $K$ is unavoidable (see section 4.1 in \cite{Potworowski2011} and references therein). We used a regularization parameter $\lambda=0.001 |e_{max}|$, where $e_{max}$ is the principal eigenvalue of $K$.

The attainment of $K$ and $\tilde{K}$ turns out to be computer-costly for high density MEAs and dense temporal sampling, but paraletization, specially on
GPUs, solves this issue.
We have performed comparison of both methods over experimental data that has
no failing channels in order to adjust the parameters of the second method. We
have obtained consistent results between them.

@@

Center of Mass Analysis (CMA) has been used previously in order to track putative  ``trajectories'' of the activity or information across neural tissue or even the whole brain \cite{Chao05, Chao07, Manjarrez07, Manjarrez09}. In such works the chosen ``mass'' was an heuristic measure adecuate for representing the corse-grain picture of displacement of activity. These present the problem of justifing the choice made on practical grounds. For calculating a vectorial average, the adecuate wheight per vector is a density, a positive quantity that measures the concentration of something per space unit, so it is only adequate that we search for a natural density in the electrophysiological experiments. The Current Source Density fits the mathematical and interpretative requirements. We shall denote it in the following as $C$ (following \cite{Potworowski2011}).  This measure is usually presented in arbitrary units, which permits us to interpret it two-fold. One, directly from its definition, as the \emph{differential divergence} of the vectorial electric current density:
\begin{equation}
  C:=\nabla \cdot J
\end{equation}
Due to conservation of charges, this is the same as as the temporal variation of density of charge:
\begin{equation}
  C=\frac{\partial \rho}{\partial t}
\end{equation}
Current Source Density as a mathematical density (a quasi-probability density function)  allows  to rigorously apply the concepts of vector averaging. This use has been overlooked. This might have occurred because  samplings at lesser spatial resolution did not permit estimations of this quantity with enough precision to make reliable spatial implications, even with the use of sophisticated CSD estimation methods. In our case, we want to use the density as a mathematician would: a function which assigns weights to points in the space of interest and permits us to carry linear operations over them.  The space here is the sampling points of the electrodes at a given instant in time, and the weight, the CSD, is amount of interchange of charged ions from the inside to the outside of cellular membranes, per unit space.  This is an indicator of activity in the neurons, therefore the CSD is also, indirectly, a \emph{``density of local activity''}.

Apparently, CSD would not fit the requirements of non-negativity for its use as a mathematical density, but this turns out to be not a problem.  CSD separates the measured activity into three distinct sets, namely, the set of all sources, the set of all sinks and the zero set. The separation is not quantitative but qualitative: even when the signs of electric charge are arbitrary, the distinction between them is not. Therefore, CSD represents on each of the two active sets (sinks and sources) exactly what we need, a unit of concentration per unit space, in this case, per area, and the sign can be ignored, as long as we perform the operations in each set separately. Also, the separation by the zero value set of the CSD is in contrast to the zero value of the LFP, where it has no precise meaning as it only represents a practical reference value. This is acknowledged in most of the literature, where only the scales of the LFP recording are shown, without reference to the sign of the values. In the CSD representation,  the so called zero set (all the points in our space which have a recorded CSD of zero)  has a precise meaning: it is an instantaneous lack of activity and a border between sinks and sources. This shall be exploited further in the section below.

@@

To clarify the concept and how we are we using it an example is in order. In figure \ref{Disjuntos} A we show geometrical sets enclosed by simple curves as color patches, where every color indicates a Set. The set $A$ consists of a single component, that means that one can make  path between any two points in the set that consist entirely of points of the same set. In contrast, the set $B$ is made of two disjoint components, labeled $B1$ and $B2$. If one chooses a point in each of the components, one cannot draw a path joining them without leaving the set. But each one of those components is connected, meaning that if we choose the pair of points in one component, then again we can trace a path joining them in the set. So the set $B$ consists of two disjoint components, but each component can be regarded as a connected set on their own. Likewise the set $C$ is made of three disjoint components. In the subfigure \ref{Disjuntos} Aiii we show the center of mass of the sets. The set $A$ is concave, so, it can have a center of mass outside the set. The
set $B$ has also a center of mass outside the set. It has two disjoint components that are aproximatelly
equall and separated by a long distance (this is not an implication: a set can be disjoint and have
a CM inside the set). For the set $C$ our aproach has been used: we treat every disjoint component of the
set as an independent entity and calculate the center of mass for each one. Notice that this doesn't preclude the posibility of having a concave subset with a CM that lies outside. But, as we shall see,
as the electrode resolution is finite, the posibility of having highly non compact shapes is unlikely, and they
tend to have a CM inside or near their borders. In the figure \ref{Disjuntos} B we show this with experimental
data (we are using evoked activity data, as explained in section \ref{sec:evocada}). In graph Bi we
sepparate the CSD data into the sink and sources set. We take sinks as example. In the graph Bii
we label each disjoint component, obtained by a single pass algorhythm. There are 25 components, and
the biggest two are not strictly convex but they have tolerable deviations from it. In the
subfigure Biii we indicate the CM of only those components that are above certain threshold of
integrated intensity (the dot is proportional in area to such intensity, there are five CM to be
seen). Each CM gets listed as a probable step in a trajectory.

\begin{figure}
  \includegraphics[width=0.9\textwidth]{Figuras/Disjuntos01.pdf}
  \caption{ Explanation of the geometrical concepts used on our analysis. In subfigure A we show
    with an artificial example the concepts of disjointness and convexivity. The subfigure Ai shows
    three sets, $A$, $B$, and $C$. $A$ fails to be convex because it is concave, and $C$ fails to be
    be convex because it is made of two disjoint components. In each of them is possible to draw a line
    connecting two points inside the set that is not entirely inside the set. In $B$ that is not possible, the set $B$ is convex in a mathematical sense. In subfigure Aii we made some modification to show that
    each set can be made of many disjoint components, and each one of them can be convex or not.  In Aiii we show the center of mass for $A$ and $B$ taken as a whole, while for $C$ we break the set into each of the disjoint components and calculate each CM separately. There is no guarantee that each disjoint component would be convex, but as the resolution of the data is finite, the smaller components are more likely to
    be convex. In figure B we shoe everything with a frame of experimental data. In Bi we sepparate the
    data into three sets, Sources, Sinks, and Zeros. We discard the Zeros and keep the interesting ones.
    In Figure Bii we take the Sink Set and separate it into disjoint components, that are labeled from one to 25 in this particular example. From this subsets we calculate the total mass of each one and only
    considerate for the next step those who are above certain threshold of integrated intensity. The
    next step is shown in Biii, where we plot the Center of Mass for each of such subsets. Deviations
    from convexity seem to be unimportant, as most of the mass is concentrated in the inner electrodes
    of each set.
  }
  \label{disjuntos}
\end{figure}

@@

We must not take out of sight the phenomena which is producing the CSD distribution: neurons have a finite size, and the effect of their action a delimited extent. Every patch that we can identify as a disjoint component of the Sources/Sink sets indicates a coherent, joint active group of units. A center of mass for each disjoint component would then indicate an average locus for this active group, a putative center of activity to which we can give a position, intensity and time. An instantaneous snapshot of such information would depict the centers of activity of an ordered structure of neurons and their relative intensities. But it is in the successive depiction of a series of such snapshots that this analysis can show interpretative power. The appearance , displacement, and disappearance of such centers could reveal very subtle details of the dynamics of neural activity. Very small, fast displacements of the Centers of mass, may be occur due to slight asynchronous between neurons which belong to the same active group. Larger displacements and  longer intervals (more than two or so firing periods) could indicate physiological connections between different groups.
